Factors affecting health sector expenditure in OECD countries

### 2025-03-14 code/analysis-code/statistical-analysis.qmd
project part3

## Quarto

@@

```{r}
# Load necessary libraries
library(dplyr)

@@

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      # For tidy regression results
library(purrr)      # For functional programming
library(kableExtra) # For advanced table formatting

# Ensure all selected variables are numeric
mydata <- mydata %>%
  mutate(across(everything(), as.numeric))

# Function to run multiple regression for a specific category and country
run_regression <- function(data, category_vars, category_name) {
  # Drop variables with all NA values within the category
  category_vars <- category_vars[sapply(data[category_vars], function(x) !all(is.na(x)))]

@@

# Function to run regression for all countries and a specific category
run_regression_all_countries <- function(category_vars, category_name) {
  results <- mydata %>%
    group_by(CountryName) %>%
    group_modify(~ {
      regression_results <- run_regression(.x, category_vars, category_name)
      if (!is.null(regression_results)) {
        return(regression_results)
      } else {
        return(tibble())  # Return empty tibble if no results
      }
    })

  return(results)
}

# Run regression for each category and create tables
economic_results <- run_regression_all_countries(variable_categories$Economic, "Economic")
health_results <- run_regression_all_countries(variable_categories$Health, "Health")
social_results <- run_regression_all_countries(variable_categories$Social, "Social")
government_results <- run_regression_all_countries(variable_categories$Government, "Government")
trade_results <- run_regression_all_countries(variable_categories$Trade, "Trade")

@@

# Remove the CountryName column
all_results <- all_results %>%
  select(-CountryName)

@@

# Create an advanced table using kableExtra
all_results %>%
  kable("html", caption = "Regression Results by Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "500px") %>%
  row_spec(which(all_results$Significant == "YES"), background = "#D9F2D9") %>%  # Highlight significant rows in green
  row_spec(which(all_results$`Predictor variables` %in% paste("Independent variables from", unique(all_results$Category), "category")), 
           bold = TRUE, background = "#FFA500") %>%  # Highlight separator rows in orange
  column_spec(1, bold = TRUE)  # Bold the "Predictor variables" column
```

>>>>

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      
library(kableExtra) 
library(webshot)
library(magick)
library(here)
library(modelsummary)
library(plm)
library(lmtest)
```

At first we will import the dataset

@@

We will now pick the most useful variables from the large pool of 1400+ variables in our dataset, and we will then categorize these variables based on the aspects they cover. For example, if there is avariable named "Annual GDP growth", we will categorize it as an "Economic" variable. Similarly, if there is a variable named "Unemployment rate" or "Total population", we will categorize them as "Social" variable.

```{r}

@@

Now that we have selected our variables, we will try to look at the relation between our dependent variable "Health Expenditure Per Capita \$USD" with other variables from a specific category.

```{r}

# Convert all columns to numeric except CountryName, CountryCode, and SeriesCode
mydata <- mydata %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

# Function to run regression for a specific category of independent variables
run_regression <- function(data, category_vars, category_name) {
  # Drop variables with all NA values within the category
  category_vars <- category_vars[sapply(data[category_vars], function(x) !all(is.na(x)))]

@@

# Run regression for each category and create tables
economic_results <- run_regression(mydata, variable_categories$Economic, "Economic")
health_results <- run_regression(mydata, variable_categories$Health, "Health")
social_results <- run_regression(mydata, variable_categories$Social, "Social")
government_results <- run_regression(mydata, variable_categories$Government, "Government")
trade_results <- run_regression(mydata, variable_categories$Trade, "Trade")

@@

@@

# Create an advanced table using kableExtra
table_html <- all_results %>%
  kable("html", caption = "Regression Results by Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "500px") %>%
  row_spec(which(all_results$Significant == "YES"), background = "#D9F2D9") %>%  # Highlight significant rows in green
  row_spec(which(all_results$`Predictor variables` %in% paste("Independent variables from", unique(all_results$Category), "category")), 
           bold = TRUE, background = "#FFA500") %>%  # Highlight separator rows in orange
  column_spec(1, bold = TRUE)  # Bold the "Predictor variables" column

# Display the table
table_html
```

In the above table, we can observe that a lot of factors affect the dependent variable and a lot of them don't. We can also observe that a lot of our variables have missing values make it unavailable for the regression estimation.

Now, we will save our regression result is a HTML table so that we can use to present it in our manuscript.

```{r}
# Save the table as a PDF file
html_file <- here("results", "tables", "regression_results.html")
save_kable(table_html, file = html_file)

```

Now that we have tested some variables in our model across different categories, we will pick the most useful variable and add them into our preliminary model. We will also add other variable that were not present in the above table. Variable selection is a continuous process and the variables selected in the preliminary model may also be dropped later to make space for new variables that may appear as important.

```{r}

model <- lm(Current_health_expenditure_per_capita_current_US ~ 
              GDP_growth_annual_ + 
              Income_share_held_by_lowest_20 + 
              Control_of_Corruption_Estimate + 
              Life_expectancy_at_birth_total_years + 
Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
              Literacy_rate_adult_total__of_people_ages_15_and_above + 
              Trade__of_GDP+ Foreign_direct_investment_net_inflows_BoP_current_US+ 
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population
  , 
            data = mydata)

# Save the regression results to an HTML file
modelsummary(
  model,
  output = here("results", "tables", "preliminary_regression_results.html"),
  title = "First Attempt at Regression Model",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  ),
  encoding = "UTF-8"  # Ensure UTF-8 encoding
)
```

The variables have a significant impact on the dependent variable as expected. But, as this is a time series dataset, we must also test for autocorrelation. There are many other tests as well which can be performed on time-series models such ADF test for unit root, and complex models such as ARCH and GARCH can also be used. However, we will only perform simple procedures and try to build a simple model.

# Autocorrelation test

Below, we will perform the autocorrelation test.

```{r}
# Perform Durbin-Watson test
dw_test <- dwtest(model)
print(dw_test)
```

Unfortunately, the Durbin-Watson test is significant, and our model shows presence of autocorrelation.

We will try to use a lagged model to solve the issue of autocorrelation.

```{r}
# Create a lagged version of the dependent variable
mydata$Current_health_expenditure_per_capita_current_US_lag1 <- lag(mydata$Current_health_expenditure_per_capita_current_US, 1)

mydata$GDP_growth_annual_lag1 <- lag(mydata$GDP_growth_annual_, 1)

# Refit the model with lagged variables
model_lagged <- lm(Current_health_expenditure_per_capita_current_US ~ 
                     GDP_growth_annual_ + GDP_growth_annual_lag1 +
                     Income_share_held_by_lowest_20 + 
                     Control_of_Corruption_Estimate + 
                     Life_expectancy_at_birth_total_years + 
                     Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
                     Literacy_rate_adult_total__of_people_ages_15_and_above + 
                     Trade__of_GDP + 
                     Foreign_direct_investment_net_inflows_BoP_current_US + 
                     Access_to_electricity_rural__of_rural_population +
                     Age_dependency_ratio__of_workingage_population +
                     Current_health_expenditure_per_capita_current_US_lag1,  # Include lagged dependent variable
                   data = mydata)

# autocorrelation test

dw_test_lagged <- dwtest(model_lagged)
print(dw_test_lagged)
```

Unfortunately, the DW test is still significant, and lagged model also shows auto correlation.

We will try to use a country fixed effect to account for the all time-invariant differences between countries (e.g., cultural, institutional, or geographical factors). Although these country fixed effect is not a solution to the autocorrelation problem in itself, we believe that the new model specification will show a different property than our previous two models with ordinary pooled OLS and lagged model.

```{r}
# Fit a fixed effects model using lm() with dummy variables
fixed_model_lm <- lm(Current_health_expenditure_per_capita_current_US ~ 
                       GDP_growth_annual_ + 
                       Income_share_held_by_lowest_20 + 
                       Control_of_Corruption_Estimate + 
                       Life_expectancy_at_birth_total_years + 
   Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate +
                       Literacy_rate_adult_total__of_people_ages_15_and_above + 
                       Trade__of_GDP + 
                       Foreign_direct_investment_net_inflows_BoP_current_US + 
                       Access_to_electricity_rural__of_rural_population +
                       Age_dependency_ratio__of_workingage_population + 
                       factor(CountryName),  # Include dummy variables for each entity
                     data = mydata)

#autocorrelation test

dw_test_lagged <- dwtest(fixed_model_lm)
print(dw_test_lagged)
```

Thankfully, our country fixed effect model no longer shows autocorrelation. We will now be able to build upon this model for our further analysis.

### 2025-03-14 products/manuscript/Manuscript.qmd
project part3

This uses HTML as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like docx or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

@@

* Prasanga Paudel$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)

@@

$\land$ Corresponding author: some@email.com

@@

# Summary/Abstract
_The data is based on World Development Indicator from World Bank Group. This is a comprehensive database that includes information including employment, public and private expenditure in health sector, fertility, immunization, literacy, mortality, and many more.The dataset has a total of 1496 variables and across 11 years from 2010 to 2020. The actual source of data has information from 1960 to 2023 but to deal with the problem of missing values in recent years and overcomplication in analysis due to too many year data , we will focus on 10 recent years between 2010 and 2020._

@@

## General Background Information
_We will try to analyze the most influential factors affecting the expenditure in health sector by different countries. The predictor variables will include factors ranging from education, mortality, population distribution among rural and urban spaces, and so on. We would use multiple regression analysis with health sector expenditure as dependent variable and the variable of interests as predictors for this panel data._

_We are using data from OECD members for analysis, and there have already been some research which have focused on GDP as predictor for health sector expenditure. Some research have also focused on the relation between life expectancy and health expenditure. However, these research were based on fixed effect and random effect models. In our analysis we will try to follow OLS multiple regression model._

## Description of data and data source
_This is a time series data based on 38 member countries from the Organisation for Economic Co-operation and Development (OECD). This organization is a unique forum with governments from 37 democracies collaborate to develop policy standards to promote sustainable economic growth. However, the data obtained from this dataset is crude and needs to be treated before attempting analysis based on it. For example, in its present form, the variable name is a long sentense describing the data it contains.The years are listed as column names and variables are listed as Series._

@@

_The main objective of this research is to:_

_1. To analyze the factors that affect the expenditure in health sector among OECD countries._

_2. To analyze the magnitude of the influence of these factors on health expenditure_

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

@@

## Data aquisition
_The data for this research is available at the official World Bank- Data Bank website.It is a free database managed by world bank group and can be accessed anytime. We can apply various filters as per our need and export the data as a .CSV or an .xlsx file._

@@

_This was the state of the data when first downloaded_
```{r}
head(data)
```

@@

_This is the present state of the dataset_

@@

_Here, we will now use a combination of tables and figues to have a look into the data._

@@

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of variables"
#| echo: FALSE
table_html <- readRDS(here::here("results", "tables", "health_exp_2010.rds"))
table_html

@@

@fig-health-exp-aus shows the distribution of "Health Expenditure as a Percentage of GDP changes over time for Australia.

```{r}
#| label: fig-health-exp-aus
#| fig-cap: "Health Expenditure (% of GDP), Country: Australia"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "health_exp_aus.png"))
```

@fig-pop-growth-aus shows the distribution  of Health Expenditure as a Percentage of GDP over the Years Comparing with shows any reation with annual population growth for Australia.This figure shows that it has very less in common with @fig-health-exp-aus as long as distribution and correlation is concerned.

```{r}
#| label: fig-pop-growth-aus
#| fig-cap: "Health Expenditure (% of GDP), Country: Australia"
#| echo: FALSE
knitr::include_graphics(here("results","figures","pop_growth_aus.png"))
```

@fig-he_per_capita shows the distribution of Health Expenditure Per Capita (USD) over the Years. It shows that it is some what related to @fig-pop-growth-aus in it's distribution. So we can say that there is indeed a presence of some pattern in the variables of interest.

```{r}
#| label: fig-he_per_capita
#| fig-cap: "Health Expenditure Per Capita (USD, Country: Australia"
#| echo: FALSE
knitr::include_graphics(here("results","figures","health_exp_per_capita_aus.png"))
```
 I think it will be better for us to use the variable "Health Expenditure Per Capita $USD" as a dependent variable in our analysis as it show more variation and connections with other variables.

## Basic statistical analysis

_To be added_

@@

_To be added._

@@

## Summary and Interpretation
_To be added._

## Strengths and Limitations
_To be added._

## Conclusions
_To be added._

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

>>>>

This uses HTML as output format.

@@

-   Prasanga Paudel$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)

@@

$\land$ Corresponding author: some\@email.com

@@

# Summary/Abstract

The data is based on World Development Indicator from World Bank Group. This is a comprehensive database that includes information including employment, public and private expenditure in health sector, fertility, immunization, literacy, mortality, and many more.The dataset has a total of 1496 variables and across 11 years from 2010 to 2020. The actual source of data has information from 1960 to 2023 but to deal with the problem of missing values in recent years and overcomplication in analysis due to too many year data , we will focus on 10 recent years between 2010 and 2020.

@@

## General Background Information

We will try to analyze the most influential factors affecting the expenditure in health sector by different countries. The predictor variables will include factors ranging from education, mortality, population distribution among rural and urban spaces, and so on. We would use multiple regression analysis with health sector expenditure as dependent variable and the variable of interests as predictors for this panel data.

We are using data from OECD members for analysis, and there have already been some research which have focused on GDP as predictor for health sector expenditure. Some research have also focused on the relation between life expectancy and health expenditure. However, these research were based on fixed effect and random effect models. In our analysis we will try to follow OLS multiple regression model.

## Description of data and data source

This is a time series data based on 38 member countries from the Organisation for Economic Co-operation and Development (OECD). This organization is a unique forum with governments from 37 democracies collaborate to develop policy standards to promote sustainable economic growth. However, the data obtained from this dataset is crude and needs to be treated before attempting analysis based on it. For example, in its present form, the variable name is a long sentense describing the data it contains.The years are listed as column names and variables are listed as Series.

@@

*The main objective of this research is to:*

*1. To analyze the factors that affect the expenditure in health sector among OECD countries.*

*2. To analyze the magnitude of the influence of these factors on health expenditure*

@@

## Data aquisition

The data for this research is available at the official World Bank- Data Bank website.It is a free database managed by world bank group and can be accessed anytime. We can apply various filters as per our need and export the data as a .CSV or an .xlsx file.

@@

This was the state of the data when first downloaded

```{r}
head(data)
```

@@

This is the present state of the dataset

@@

*Here, we will now use a combination of tables and figues to have a look into the data.*

@@

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of variables"
#| echo: FALSE
table_html <- readRDS(here::here("results", "tables", "health_exp_2010.rds"))
table_html
```

@@

But, before moving on to be fixed at a single variable as our dependent variable, we will also look if there are any alternative dependent varibles that could prove to be more useful.

@fig-health-exp-aus shows the distribution of "Health Expenditure as a Percentage of GDP changes over time for Australia. Here we can see that there is not much variation in the data. Therefore we will also look at other similar variables.

```{r}
#| label: fig-health-exp-aus
#| fig-cap: "Health Expenditure (% of GDP), Country: Australia"
#| echo: FALSE 
#| out.width: "60%"
knitr::include_graphics(here("results", "figures", "health_exp_aus.png"))
```

@fig-he_per_capita shows the distribution of Health Expenditure Per Capita (USD) over the Years. It can be observed that this variable indeed has variation and could be useful as our dependent variable.

```{r}
#| label: fig-he_per_capita
#| fig-cap: "Health Expenditure Per Capita (USD, Country: Australia"
#| echo: FALSE
#| out.width: "60%"
knitr::include_graphics(here("results","figures","health_exp_per_capita_aus.png"))
```

@fig-pop-growth-aus shows the distribution of population growth of australia.This figure shows that it has very less in common with @fig-health-exp-aus and more in common with @fig-he_per_capita

```{r}
#| label: fig-pop-growth-aus
#| fig-cap: "Population Growth, Country: Australia"
#| echo: FALSE
#| out.width: "60%"
knitr::include_graphics(here("results","figures","pop_growth_aus.png"))
```

I think it will be better for us to use the variable "Health Expenditure Per Capita \$USD" as a dependent variable in our analysis as it show more variation and connections with other variables.

# Basic statistical analysis

## Preparing a model

To help finalize our model, we will first try to look at the relation between our dependent variable "Health Expenditure Per Capita \$USD" with other variables from a specific category. Specifically, we will pick the most useful variables from the large pool of 1400+ variables in our dataset, and we will then categorize these variables based on the aspects they cover. For example, if there is avariable named "Annual GDP growth", we will categorize it as an "Economic" variable. Similarly, if there is a variable named "Unemployment rate" or "Total population", we will categorize them as "Social" variable

We will then run our initial regression with "Health Expenditure Per Capita \$USD" as dependent variable, and variables from each of the category as independent variables in separate regressions

@fig-regression-results shows the different regresion results based on different categories. The categories are: Economic, Health, Social, Government, Trade. We will pick the most useful variables among the variables that came up significant and create our preliminary model for our research. However, this would not be the end of our variable selection process.

```{r}
#| label: fig-regression-results
#| fig-cap: "Regression Results by Category"
#| echo: FALSE
htmltools::tags$iframe(
  src = here("results", "tables", "regression_results.html"),
  width = "100%",
  height = "600px"
)
```

## Observing the relationship with dependent variable

@fig-correlation-p1 show the correlation of different variables that had a statistical significant effect in @fig-regression-results (and other variables of interest). We are just trying to see the pattern present in the dataset for the year 2010, with all the countries treated as cross-sectional units.

```{r}
#| label: fig-correlation-p1
#| fig-cap: "Correlation based on 2010 data"
#| echo: FALSE
#| out.width: "100%"
knitr::include_graphics(here("results","figures","correlation_plot1.png"))
```

We do the same with few other variables.

```{r}
#| label: fig-correlation-p2
#| fig-cap: "Correlation based on 2010 data"
#| echo: FALSE
#| out.width: "100%"
knitr::include_graphics(here("results","figures","correlation_plot2.png"))
```

It can be seen that most of the variables seem to have a high correlation with our variable of interest. Although some variables lack observations, we must recall that this is just for the year 2010, and the whole dataset might have a different distribution pattern.

## Proposed Model for the research

Based on the informations obtained in @fig-regression-results, we have selected variable that would best explain our dependent variable. The model is as follow:

Current Health Expenditure Per Capita (\$USD) = $b_0$ + $b_1$ Annual GDP growth + $b_2$ Income share held by lowest 20% + $b_3$ Control of corruption + $b_4$ Life expectancy at birth + $b_5$ Unemployed youth population as a percentage of total labor force + $b_6$ Adult literacy rate + $b_7$ Trade as a percentage of GDP + $b_8$ FDI net inflow + $b_9$ Access to electricity (rural) + $b_{10}$ Age dependency ratio

```{r}
#| label: fig-preliminary-regression-results
#| fig-cap: "Preliminary Regression Results"
#| echo: FALSE
#| out.width: "70%"
knitr::include_graphics(here("results", "tables", "preliminary_regression_results.png"))

But, it turns out that the model shows significant amount of autocorrelation

```{r}
#| label: fig-DW-test-results
#| fig-cap: "Durbin Watson Test Results"
#| echo: FALSE
#| out.width: "25%"
knitr::include_graphics(here("results", "figures", "durbin_watson_test_results.png"))

**Therefore, we will use a regression model with country fixed effect which is given below:**

Current Health Expenditure Per Capita (\$USD) = $b_0$ + $b_1$ Annual GDP growth + $b_2$ Income share held by lowest 20% + $b_3$ Control of corruption + $b_4$ Life expectancy at birth + $b_5$ Unemployed youth population as a percentage of total labor force + $b_6$ Adult literacy rate + $b_7$ Trade as a percentage of GDP + $b_8$ FDI net inflow + $b_9$ Access to electricity (rural) + $b_{10}$ Age dependency ratio + country fixed effect.

```{r}
#| label: fig-DW-test-results-fe
#| fig-cap: "Durbin Watson Test Results FE-Model"
#| echo: FALSE
#| out.width: "30%"
knitr::include_graphics(here("results", "figures", "durbin_watson_test_results_fe.png"))

@@

*To be added.*

@@

## Summary and Interpretation

*To be added.*

## Strengths and Limitations

*To be added.*

## Conclusions

*To be added.*
